---
<<<<<<< Updated upstream
title: "Day 1 - Prac 3 - Islands"
author: "Alex Skeels and Oskar Hagen"
=======
title: "Diversification in a Model Island System"
author: "Alex Skeels and Oskar Hagen"
execute: 
  eval: FALSE
>>>>>>> Stashed changes
---

This practical uses a theoretical archipelago system. Each site (1x1 km), features temperature (i.e. mean, minimum and maximum temperature) and has a landscape structure that is generated by approximating topography, uplift dynamics, and a lapse rate of temperature with elevation. Additionally, it incorporates global temperature and sea level changes dating back to the past 5 million years (Ma).

<<<<<<< Updated upstream
You can look at an animation of the landscape [here](https://raw.githubusercontent.com/ohagen/DispDiv3/main/figures/AnimationS1_patch_dynamics.gif). We'll also have a look now.
=======
First make sure our packages are loaded.

```{r}
library(gen3sis)
library(terra)
```

You can look at an animation of the landscape [here](https://raw.githubusercontent.com/ohagen/DispDiv3/main/figures/AnimationS1_patch_dynamics.gif). We'll also have a quick look now.
>>>>>>> Stashed changes

# Dynamic Island Landscape

```{r}
<<<<<<< Updated upstream
# load landscapes

lc <- readRDS(file.path("data", "landscapes", "islands","landscapes.rds"))

# class(lc) "list"

# get names of landscape variables

=======
# libraries
library(gen3sis)
library(terra)

# load landscapes
lc <- readRDS(file.path("data", "landscapes", "islands","landscapes.rds"))


# get names of landscape variables
>>>>>>> Stashed changes
names(lc)

# get first time step

first_step_pos <- ncol(lc$mean_temp)

# get first 10 sites of mean temperature for the 2 last time steps and the first (oldest) time step

lc$mean_temp[100:110, c(1:4, first_step_pos)]

# plot mean_temp for first and last time step

plot(rast(lc$mean_temp[ ,c(1:3, first_step_pos)]))
```

<<<<<<< Updated upstream
Plot out the landscape to understand it's dynamics. First define the function.

```{r}
plot_landenv <-  function(df, times=100:0, reverse=F, speed=0.1){
  # df is a data frame with x, y coordinates and time steps as columns with environmental variables
  # reverse is a boolean. Reverse the order of time steps?
  # times is either a vector of time steps or a character "all"
  # speed in seconds
  #df <- lc$temp
  if (times[1]=="all"){
   times <- names(df)[!names(df)%in%c("x", "y")] 
  }
  if (reverse){
    times <- rev(times)
  }
  for (ti in times){
    ri <- rast(df[,c("x", "y", ti)], type="xyz")
    plot(ri, main=paste(as.numeric(ti)/100, "Ma"))
    Sys.sleep(speed)
  }
}
```

Then run the animation function

```{r}
plot_landenv(lc$elevation, times=seq(from=1, to=500, by=50), speed=0.1)
```

load in a config

```{r}
# load config
cf <- create_input_config(config_file = "configs/island_config_simple.R")
# list all main elements of the config file
names(cf$gen3sis)
# list all elements of the general section, i.e. the main settings and not so much on the eco-evolutionary processes
names(cf$gen3sis$general)
```

# Run a basic simulation

```{r}
# run simulation
sim <- run_simulation(config = here("config/config_simple.R"), 
                      landscape = here("space"), 
                      output_directory = here("output"))
```

```{r}
sim <- readRDS(here("output/config_simple/sgen3sis.rds"))
=======
Look to see if there are any temporal trends in temperature or elevation.

```{r}
# plot the change in mean tempereature through time
mean_temperatures <- colMeans(lc$mean_temp[, 3:ncol(lc$mean_temp)], na.rm=T)
plot(0:500,mean_temperatures ,  xlab="time", ylab="mean temperature")

# plot the change in mean elevation through time
mean_elevation <- colMeans(lc$elevation[, 3:ncol(lc$elevation)], na.rm=T)
plot(0:500,mean_elevation ,  xlab="time", ylab="mean elevation")

```

# Config Time! - Model 1

load in a the m1 config

```{r}
# load config
conf_m1 <- create_input_config("configs/config_islands_simple_Day1Prac3_M1.R")

# list all main elements of the config file
names(conf_m1$gen3sis)

# list all elements of the general section, i.e. the main settings and not so much on the eco-evolutionary processes
names(conf_m1$gen3sis$general)

# if you're game take a look at the whole thing - whats different to the South America example?
conf
```

Ok now run the sim.

```{r}
# run simulation
sim_m1 <- run_simulation(config = "configs/config_islands_simple_Day1Prac3_M1.R", 
                      landscape = "data/landscapes/islands", 
                      output_directory = "output/islands")
```

Invetsigate the output

```{r}
sim_m1 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M1/sgen3sis.rds")

>>>>>>> Stashed changes
```

```{r}
#check elements inside the sim object
<<<<<<< Updated upstream
names(sim)

# visualize the outputs
plot_summary(sim)

# plot richness from summary in custom fashion
na_mask <- is.na(lc$elevation[,"0"])
rich <- sim$summary$`richness-final`
rich[na_mask,3] <- NA
plot(rast(rich, type="xyz"), col=c("grey", gen3sis::color_richness_non_CVDCBP(max(rich, na.rm=T))), main="Richness")


# plot richness at time step 32 using saved data
sps32 <- readRDS(here("output/config_simple/species/species_t_32.rds"))
lc32 <- readRDS(here("output/config_simple/landscapes/landscape_t_32.rds"))
plot_richness(sps32, lc32)
```

```{r}
phy <- read.nexus(file.path(here("output/config_simple/phy.nex")))
plot(phy)
```

# Customize simulations

For an example of a more complex simulation, with species abundances, traits trading off and evolving check the config_M2_TH.R.

We will modify it so that allopatric speciation is faster.

\`\`\`{r, eval=FALSE}

conf \<- create_input_config(here("config/config_M2_TH.R"))

conf\$gen3sis\$speciation\$divergence_threshold = 30

\`\`\`

We will also change the observer function to save the presence/absence matrix for each time step.

\`\`\`{r, eval=FALSE}

conf\$gen3sis\$general\$end_of_timestep_observer = function(data, vars, config){

plot_richness(data\$all_species, data\$landscape)

\# make p/a matrices

out_dir \<- config\$directories\$output

if(!file.exists(file.path(out_dir,"occs"))){

dir.create(file.path(out_dir, "occs"))

}

\# cell names

all_cells \<- rownames(data\$landscape\$coordinates)

\# get 0 for absence and 1 for presence in each grid cell

asp \<- do.call(cbind,

lapply(data\$all_species, FUN = function(x) {

ifelse(all_cells %in% names(x\$abundance), 1, 0)

}))

\# colnames are species names

colnames(asp ) \<- unlist(lapply(data\$all_species, function(x){x\$id}))

\# column bind with x/y coordinates

presence_absence_matrix \<- cbind(data\$landscape\$coordinates, asp)

saveRDS(presence_absence_matrix,

file=file.path(out_dir,"occs", paste0("pa_t\_",vars\$ti, ".rds")))

}

\`\`\`

\`\`\`{r, eval=FALSE}

simod \<- run_simulation(config=conf,

landscape=here("space"),

output_directory=tempdir())

\`\`\`

The dynamics is different from the simple simulation, with more species and a some extinction events.

\`\`\`{r, echo=FALSE}

plot_summary(readRDS(here("output/config_M2_TH_mod/sgen3sis.rds")))

plot(read.nexus(file.path(here("output/config_M2_TH_mod/phy.nex"))))

\`\`\`

\<div style="background-color: lightgreen; padding: 1px;"\>

🏋💻 \*\*Exercise\*\* \[15 min\]

Review the config_M2_TH.R file and try to understand what it's doing. Consider how you might modify the configuration or apply it to a specific research question.

\</div\>

\## 5. Troubleshoot

Creating or modifying a gen3sis configuration can definitely lead to some weird errors, especially since it's so flexible. That's the downside, along with the steep learning curve. But if you're not too overwhelmed, you're doing great!

Here are some handy debugging tips for when you run into those pesky errors:

browser(): This function lets you pause the execution and explore what's going on. It's like a pit stop where you can check out variables and step through the code.

To make your R session enter browser mode whenever you hit an error, you can use: \*options(error = recover)\*. This will help you diagnose and fix issues more easily.

You can also condition browser calls, which is very useful when you want to stop execution at a specific time step or when a certain condition is met. Here's an example:

\`\`\`{r, eval=FALSE}

\# Use 'stop_time' to halt execution at a specific timestep in the landscape object:

stop_time \<- "62"

get_dispersal_values \<- function(n, species, landscape, config) {

if (landscape\$timestep == stop_time) {

browser()

}

\# You can also check the 'vars' object for the current timestep:

vars \<- dynGet("vars", inherits = TRUE)

if (vars\$ti == stop_time) {

browser()

}

}

\`\`\`
=======
names(sim_m1)

# visualize the outputs
plot_summary(sim_m1)

# plot richness from summary in custom fashion
na_mask <- is.na(lc$elevation[,"0"])
rich_m1 <- sim_m1$summary$`richness-final`
rich_m1[na_mask,3] <- NA
plot(rast(rich_m1, type="xyz"), main="Richness")
```

Pick a couple of random time points to look at specifically.

```{r}
# plot richness at time step 32 using saved data
sps32 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M1/species/species_t_32.rds")
lc32 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M1/landscapes/landscape_t_32.rds")
plot_richness(sps32, lc32)

# plot richness at time step 12 using saved data
sps12 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M1/species/species_t_12.rds")
lc12 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M1/landscapes/landscape_t_12.rds")
plot_richness(sps12, lc12)

```

Now look at the phylogeny

```{r}
# Now lets look at the Phylogeny
library(ape)
phy_m1 <- read.nexus("output/islands/config_islands_simple_Day1Prac3_M1/phy.nex")
plot(phy_m1)
```

# Customize simulations - M2

Lets start changing features of the config and compare the outputs to our first model.

```{r}
# load in config M2

conf_m2 <- create_input_config("configs/config_islands_simple_Day1Prac3_M2.R")

# currently populations need 10 time steps to complete speciation

conf_m1$gen3sis$speciation$divergence_threshold

# now populations only need 5 time steps to complete speciation

conf_m2$gen3sis$speciation$divergence_threshold

#We have also change the observer function to save the presence/absence matrix for each time step.
# see original config
conf_m1$gen3sis$general$end_of_timestep_observer 

# see modeified config
conf_m2$gen3sis$general$end_of_timestep_observer 
```

Run M2

```{r}
# run the new model
sim_m2 <- run_simulation(config="configs/config_islands_simple_Day1Prac3_M2.R", landscape="data/landscapes/islands", output_directory="output/islands")

# From the small tweak to the divergence factor we can see more species being generated

# original dynamics
plot_summary(sim_m1)

# modified dynamics
sim_m2 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M2/sgen3sis.rds")
plot_summary(sim_m2)

```

## Exercise

Review the M2 file and try to understand what it's doing. How are the dynamics different to M1, and how are they similar? Consider how you might modify the configuration or apply it to a specific research question.

# Go cray - Model 3

The differences between M1 and M2 are predictable. Speciation happens a bit faster in M2, and you end up with more species. Overall, however, the patterns of richness are pretty similar. This suggests that this change in the rate of speciation isn't having a big impact on geographic variation in where species are forming (at least from a glance). So let's try a new model that has a few more tweaks to see a more measurable effect.

```{r}
#load in the config for model M3
conf_m3 <- create_input_config("configs/config_islands_simple_Day1Prac3_M3.R")

# take a look at the initialisation
conf_m3$gen3sis$initialization

```

There is a big differences here. In the iniotialisation stage we start with 4 species - one on each island but this time we give each species a different value of it's dispersal ability and temperature width trait (niche breadth), and these are inversely related - high dispersing species have narrow niche breadths and low dispering species have wide niche breadths. This may create an interesting tradeoff in dynamics of island colonization.

The evolution (mutation) function is also different. Here instead of a process of homogenization followed by random drift, we are implementing zn approximation of an Ornstein-Uhlenbeck model of trait evolution. This is model where traits of each geographic cluster evolution towards the optimal value for the trait based on what the average temperature is across the population's range. Basically, its a more adaptive model than before.

Run the simulation

```{r}
sim_m3 <- run_simulation(config="configs/config_islands_simple_Day1Prac3_M3.R", 
                         landscape="data/landscapes/islands",
                         output_directory="output/islands")

```

Now compare the dynamics

```{r}
sim_m3 <- readRDS("output/islands/config_islands_simple_Day1Prac3_M3/sgen3sis.rds")

# compare the plot summaries
plot_summary(sim_m1)
plot_summary(sim_m2)
plot_summary(sim_m3)

# lets also compare the phylogenies
phy_m1 <- read.nexus("output/islands/config_islands_simple_Day1Prac3_M1/phy.nex")
plot(phy_m1)
phy_m2 <- read.nexus("output/islands/config_islands_simple_Day1Prac3_M2/phy.nex")
plot(phy_m2)
phy_m3 <- read.nexus("output/islands/config_islands_simple_Day1Prac3_M3/phy.nex")
plot(phy_m3)
```
>>>>>>> Stashed changes
