---
title: "ðŸŒ‹ Creating Worlds"
author: "Alex Skeels and Oskar Hagen"
execute: 
  eval: TRUE
---

# Introduction {.unnumbered}

Welcome, brave explorer of virtual worlds! Today, you'll learn how to create entire landscapes (no pressure!) using R. We'll shrink the universe to an **experimental size**, start with one island and gradually add more complexity; from rugosity to temperature, and even sea level changes.

In this tutorial, we will:

1.  **Create** a spatial grid (our world map).
2.  **Add** random elevation and temperature (because flat worlds are boring).
3.  **Visualize** your masterpiece (watch your creation come to life).
4.  **Modify** the shape and dynamics of your island (why stop here?).
5.  **Prepare** your world for Gen3sis (ready for simulation?).
6.  **Run** a simulation (let there be life!).

## Step 1: Setup (Gather Your Tools) {.unnumbered}

Before we build our world, let's make sure we have the right tools. No one builds a universe without a hammer... and a bunch of other R-packages.

```{r warning=FALSE, message=FALSE}
# Load the necessary libraries for plotting and data wrangling
lib <- c("terra", "raster", "lattice", "rayimage", "viridis", "gen3sis", "ggplot2", "persp", "here")
sapply(lib, require, character.only = TRUE, quietly = TRUE, warn.conflicts = TRUE)
```

## Step 2: Declare your functions (Make your tools) {.unnumbered}

Now that our libraries are ready, we need tools to help us generate the world. Below are custom functions to plot the terrain (plot_persp), simulate shape (drop and gaussian_2d), and generate random landscapes (random_landscape). These functions allow us to manipulate and visualize spatial grids with elevation data.

```{r}
plot_persp <- function(x, y, z, lcol=topo.colors, ...) {
  zz <- (z[-1,-1] + z[-1,-ncol(z)] + z[-nrow(z),-1] + z[-nrow(z),-ncol(z)])/4
  breaks <- hist(zz, plot=FALSE)$breaks
  cols <- lcol(length(breaks)-1)
  zzz <- cut(zz, breaks=breaks, labels=cols)
  persp(x, y, z, col=as.character(zzz),
        phi=30, theta=-25, ltheta = -70,
        expand = 0.5, border = NA, box = FALSE, shade = 0.75, ...)
  # list(breaks=breaks)
}


# 'shape

drop <- function(x,y, a=3, ...) { 
  r <- sqrt(x^2+y^2)
  # avoid zero division
  r[r==0] <- 1
  z <- a * sin(r)/r 
}

gaussian_2d <- function(x,y, x0, y0, sdx, sdy, a) { 
  r <- (((x-x0)^2)/(2*sdx^2))+(((y-y0)^2)/(2*sdy^2))
  z <- a*exp(-r)
  return(z)
}

# 'generator

random_lanscape <- function(x,y, n=2){
  zf <- 0
  for (i in 1:n){
  zf <- zf+outer(x, y, gaussian_2d, x0=sample(x,1), y0=sample(y,1), sdx=rnorm(1,2,5), sdy=rnorm(1,2,5), a=rnorm(1,2,2))
  }
  return(zf)
}

set_landscape_t <- function(x=seq(-10, 10, length=60),y=NA, t=-4:0, 
                            x0t=function(t){rep(0,length(t))}, 
                            y0t=function(t){rep(0,length(t))}, 
                            sdxt=function(t){2/(abs(t)+1)}, sdyt=NA, 
                            at=function(t){abs(t)+1}){
  if (any(is.na(y))){y=x}
  if (any(is.na(sdyt))){sdyt=sdxt}
  l <- list()
  for (i in 1:length(t)){
    ti <- t[i]
    l[[i]] <- outer(x, y, gaussian_2d, x0=x0t(ti), y0=y0t(ti), sdx=sdxt(ti), sdy=sdyt(ti), a=at(ti))
    names(l)[i] <- paste0("x0=",x0t(ti), "|y0=",y0t(ti), "|sdx=", sdxt(ti),"|sdy=",sdyt(ti),"|a=", at(ti))
  }
  return(l)
}

```

Every world (i.e raster) needs a grid to hold it together. We will work without a coordinate system in a simple 1 unit grid. This will allow us to focus on the shape of the landscape without worrying about real-world problems (distortion).

```{r}
# size of the grid
nlg <- 1

### single simple -----
x <- seq(-20, 20, by=nlg)
y <- x
z <- outer(x, y, gaussian_2d, x0=0, y0=0, sdx=2, sdy=2, a=10)
plot_persp(x, y, z)

```

Next, let's make it funky...

The drop function can be used to create a terrain similar to an island by generating a peak that gradually descends towards the edges. This creates a radial pattern of decreasing elevation, much like a small volcanic island with surrounding lagoons. While this isn't a very realistic model (Ask Tristan), its fun ðŸ¤“.

```{r}
### single simple drop -----
z <- outer(x, y, drop, x0=20, y0=20, a=10)
plot_persp(x, y, z)
```

## Step 3: Create a dynamic landscape {.unnumbered}

Now, let's make it dynamic! Here we use time (ti) as the input to dynamically adjust the standard deviation in the Gaussian function. Our temporal unit is time-step and cells are our spacial units.

```{r}
### multiple -----
l <- list()
for (ti in 1:6){ # set the number of time steps
  l[[ti]] <- outer(x, y, gaussian_2d, x0=0, y0=0, sdx=ti, sdy=ti, a=ti^-0.8)
  # add constant sea level
  l[[ti]][l[[ti]]<0.1] <- 0.1
}
```

## Build Your Own Tools Again, forget about steps... {.unnumbered}

The process of creating and refining tools never stops in world-building. It would be nice to have a function to plot multiple landscapes no? Below is a function to plot multiple landscapes at different time steps in one visualization.

```{r}
plot_multiple_persp <- function(l, lcol = topo.colors, scol = "darkblue", legend_title = "Legend", ...) {
  # 'l' is the list of z values...
  lv <- do.call(c, l)
  breaks <- hist(lv, plot = FALSE)$breaks
  cols <- c(scol, lcol(length(breaks) - 2))
  
  # Save the original par settings
  oldpar <- par(no.readonly = TRUE)
  on.exit(par(oldpar))
  
  # Number of subplots
  times <- length(l)
  
  # Dynamically define number of rows and columns for layout based on the number of steps
  n_cols <- ceiling(sqrt(times))  # Number of columns
  n_rows <- ceiling(times / n_cols)  # Number of rows
  
  # Set the plot margins: Adjust `mar` to make the plots closer to each other
  par(mar = c(2, 2, 2, 2)*0.3)  # Reduce margins around each plot
  
  # Set the layout for the subplots (without extra row for legend)
  par(mfrow = c(n_rows, n_cols))
  
  # Loop through the list of z-matrices and create the persp plots
  for (i in seq_along(l)) {
    zi <- l[[i]]
    zz <- (zi[-1, -1] + zi[-1, -ncol(zi)] + zi[-nrow(zi), -1] + zi[-nrow(zi), -ncol(zi)]) / 4
    zzz <- cut(zz, breaks = breaks, labels = cols)
    
    # Plot each 3D surface with persp
    persp(x, y, zi,
          zlim = range(lv, na.rm = TRUE),
          col = as.character(zzz),
          phi = 30, theta = -25, ltheta = -70,
          expand = 0.5, border = NA, box = FALSE, shade = 0.25, ...)
    
    # Add titles if names are available
    if (!any(is.na(names(l)))) {
      title(names(l)[i])
    }
    
    # If it's the first plot, add a color scale (legend) on top of it
    if (i == 1) {
      # Use a small inset to add the legend (e.g., top-right corner)
      legend("topright", legend = round(breaks, 2), fill = cols, title = legend_title, cex = 0.7, inset = 0.05)
    }
  }
}

```

## See the landscape you've crafted take shape in all its 3D glory {.unnumbered}

Use your new tool to visualize the dynamic landscape you've created.

```{r}
plot_multiple_persp(l, lcol=terrain.colors)
```

## Add more islands! {.unnumbered}

It's time to create more complex terrains by adding multiple islands. We can simulate this by layering multiple Gaussian functions. In this example we will simply add the z values

```{r}
par(mfrow=c(1,3), mai=c(1,1,1,1)*0.3)
# First island
z1 <- outer(x, y, gaussian_2d, x0=-10, y0=10, sdx=4, sdy=4, a=7)
plot_persp(x, y, z1, main="z1")

# Second island
z2 <- outer(x, y, gaussian_2d, x0=10, y0=-10, sdx=5, sdy=6, a=5)
plot_persp(x, y, z2, main="z2")

# Combine the two islands
zf <- z1 + z2
plot_persp(x, y, zf, main="z1 + z2")
```

Now let's use the same concept to create a time-evolving landscape with three islands. And see how the islands will grow and change over time.

```{r}
# Create a time-evolving landscape
l <- list(z1 = zf,
          z2 = zf + outer(x, y, gaussian_2d, x0 = 0, y0 = -3, sdx = 2, sdy = 2.5, a = 3),
          z3 = zf + outer(x, y, gaussian_2d, x0 = 0, y0 = -3, sdx = 2.5, sdy = 2.8, a = 4))
plot_multiple_persp(l, lcol=terrain.colors)

# Convert the landscapes to raster bricks
rb <- brick(lapply(l, raster))
plot(rb)

# Apply sea level cuts to the landscapes
selev <- c(1, 2, 1)
ls <- l
for (zi in 1:3) {
  ms <- ls[[zi]] <= selev[zi]
  ls[[zi]][ms] <- NA
}
rb <- brick(lapply(ls, raster))
plot(rb, col = viridis(255))


```

## Random Landscapes {.unnumbered}

Now let's generate a more complex landscape with multiple peaks to add more details to our world.

```{r}
noise <- random_lanscape(x, y, 20)
# Few random peaks
plot_persp(x, y, noise, main="Noise")

```

Now we add to our previous landscape

```{r}
# add to previous landscape
plot_persp(x, y, (noise/10)+zf, main="Noise + zf")
```

Experiment with it; there's a lot of room when creativity your own world!

```{r}

plot_persp(x, y, (random_lanscape(x, y, 200)/10)+(zf), lcol=terrain.colors)

```

## Creating an Archipelago {.unnumbered}

Let's create an archipelago with multiple islands. We will use the same concept as before, but this time we will add more islands to the landscape. I.e 6, starting from the smallest to the largest island all spread evenly in a circle.

```{r}

# Set the radius of the circle and the number of points
radius <- 17  # For a circle within -20:20 range
num_points <- 6

# Generate equally spaced angles in radians
angles <- seq(0, 2*pi, length.out = num_points + 1)[-7]  # Exclude the last angle as it repeats the first

# Compute x and y coordinates based on angles (x = cos(angle) * radius, y = sin(angle) * radius)
x_values <- radius * cos(angles)
y_values <- radius * sin(angles)

# create islands using centroids as well as an island id mask for later
island_id <- 0 # island id's
elevation <- 0 # elevation


for (island_i in 1:num_points){
  # island_i <- 1
  z_i <- outer(x, y, gaussian_2d, x0=x_values[island_i], y0=y_values[island_i], sdx=2, sdy=2, a=island_i*100) # times 100 for meter
  # plot_persp(x, y, z_i, main="z_i")
  id_mask_temp <- (z_i>0.1)*island_i # this should be the lowest sealevel of all times
  island_id <- island_id+id_mask_temp
  elevation <- elevation+z_i
}
plot_persp(x,y, elevation)

```

Like before, let's add a temporal dynamics and sea level and temperature fluctuations. We will use a sine wave to simulate temperature and sea level changes over time. We will delay the sea level change by 1 time steps, as we expect the temperature to increase before the sea level rises.

```{r}
# define number of timesteps
n_step <- 35

# cycles length in timesteps
cycles_length <- 5

# delay
delay <- 1

# generate a sequence of values that oscillates between -1 and 1 (using sine function), a sine wave
oscillatory_values <- sin(seq(0, 2*pi*cycles_length, length.out = n_step+delay))
#plot(oscillatory_values, type="l")

# get the temperature and sea level values accounding for the delay
temp <- tail(oscillatory_values, n_step)
sealevel <- head(oscillatory_values, n_step)

plot(temp, type="l", col="red")
lines(sealevel, col="blue")

# scale
limts <- list(sealevel = c(1, 100), temperature = c(-2, 2))

# function to scale oscillatory values to a given range
scale_to_range <- function(values, low, high) {
  return (((values + 1) / 2) * (high - low) + low)
}

# apply the scaling to each element in the list
temp_vec <- scale_to_range(temp, limts$temperature[1], limts$temperature[2])
sealevel_vec <- scale_to_range(sealevel, limts$sealevel[1], limts$sealevel[2])
```

Now let's apply these changes over time to our landscape.

```{r}

env <- list(elevation=list(), mean_temp=list(), sd_temp=list(), island_id=list())

# temperature at absolute 0m
temp_abs <- 27

for (t_i in 1:n_step){
  # t_i <- 1
  # get matri of NA based on habitable sites (i.e. above sea level)
  matrixNA <- elevation
  matrixNA[] <- NA
  matrixNA[elevation>sealevel_vec[t_i]] <- 1
  #habitable_mask <- elevation>sealevel_vec[t_i]
  # elevation
  env$elevation[[t_i]] <- elevation*matrixNA
  # temp
  # decrese temperature by 0.65Â°C per 100 meters (mois)
  env$mean_temp[[t_i]] <- (temp_abs+temp_vec[t_i]) - (env$elevation[[t_i]]/100)*0.65

  # island id
  env$island_id[[t_i]] <- island_id*matrixNA
}

plot_multiple_persp(env$elevation, lcol=terrain.colors)


for(i in 1:length(env$mean_temp)){
  
  for(k in nrow(env$mean_temp[[i]]):1){
    env$mean_temp[[i]][k,] <- env$mean_temp[[i]][k,] - (k/5)
  }
  env$sd_temp[[i]] <- env$mean_temp[[i]]
  env$sd_temp[[i]][] <- NA
  for(k in nrow(env$sd_temp[[i]]):1){
    env$sd_temp[[i]][k,] <- k
  }
 env$sd_temp[[i]][which(is.na(env$elevation[[i]]))] <- NA
 env$sd_temp[[i]] <- (env$sd_temp[[i]] + env$elevation[[i]]/20)/2

}


par(mfrow=c(1,3))
plot(rast(env$sd_temp[[i]]), main="temp seasonaility")
plot(rast(env$mean_temp[[i]]), main="temp mean")
plot(rast(env$elevation[[i]]), main="elevation")

# Convert the landscapes to raster bricks
rball <- lapply(env, function(xa){
  revx <- rev(xa) # rever to fit into data format... happy to hear your feedback here
  lapply(revx, function(xaxa){
    raster(xaxa, xmn=min(x)-0.5, xmx=max(x)+0.5, ymn=min(y)-0.5, ymx=max(y)+0.5)
  })
})




```



## Prepare gen3sis input {.unnumbered}

Okay, our landscape is almost ready to start getting some life on it! We will now prepare the gen3sis input using create_input_landscape, and run a simulation. We will use a config we prepared so that you can use it as a start and figure out what it is doing.

```{r, eval=FALSE}

# define cost function, crossing water as 4x as land sites
cost_function <- function(source, habitable_src, dest, habitable_dest) {
  if(!all(habitable_src, habitable_dest)) {
    return(4)
  } else {
    return(1)
  }
}


# create a gen3sis input
create_input_landscape(
  landscapes = rball,
  cost_function = cost_function,
  output_directory = file.path(here("data/landscapes/six_seasonal_islands")),
  directions = 8, # surrounding sites for each site
  timesteps = paste0((n_step-1):0, "step"),
  calculate_full_distance_matrices = TRUE,
  verbose = T,
  overwrite_output = T) # full distance matrix


exp1 <- run_simulation(
  config = file.path("configs", "config_islands_simple_Day2Prac5.R"),
  landscape = file.path(here("data/landscapes/myworld")),
  output_directory = file.path(here("output/outputs")),
  verbose = 1
)
```

## Conclusion {.unnumbered}

Congratulations! You've successfully created your own world, shaped its terrain, and even made life thrive. Your world may not be ready for civilization just yet, but don't worry---Rome wasn't built in a day either. Keep experimenting and refining your creation. Well done, master world-builder! ðŸŒâœ¨
